Testing for measurement invariance with respect to an ordinal variable

### 2012-10-16 ordinal/ordinal.Rnw
small touch-ups, removing most readme/fixme improved cite/citeNP/citeA usage, avoiding spelling out author names by hand consistency changes in refs.bib (especially title case)
to do: described m2.2 instead of m3



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@171 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

This section contains background on the theory underlying the
proposed statistics; for a more detailed
account, see Merkle and Zeileis \citeyear{MerZei12}.

@@

As mentioned previously, when $V$ is categorical with a
relatively-small number of categories, 
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models usually cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals and then
examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test and are
discussed in detail by Merkle and Zeileis \citeyear{MerZei12}. Here,
we review the tests' properties that are relevant for extending them
to the ordinal case.

@@

Test statistics that exhibit better performance in such situations 
aggregate information across parameters and possibly also across
individuals.  These test statistics include
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
with the former being a Cram{\'e}r-von Mises statistic and the latter
corresponding to a ``maximum'' Lagrange multiplier test, where the
maximum is taken across all possible divisions of individuals into two
groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
scaled by the asymptotic variance $t (1 - t)$ of the process
${\bm B}(t, \hat {\bm \theta})$.  In simulations, Merkle and Zeileis
\citeyear{MerZei12} found 
that both tests perform well when assessing simultaneous changes
in multiple factor analysis parameters, with the
$\mathit{CvM}$ test being somewhat advantageous in their particular
simulation setup. These simulations included 
situations in which subsets of model parameters were tested; such
situations are handled 
by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
that correspond to the parameters of interest.

@@

For the $\mathit{WDM}_o$ statistic, the resulting asymptotic distribution is
$\max_{j = 1, \dots, k} \max_{\ell = 1, \dots, m-1} {\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$.
Note that the effect of the outer maximum can be easily captured by a Bonferroni
correction as the $k$~components of the Brownian bridge are asymptotically independent.
Moreover, the inner maximum is taken over $m-1$ variables
${\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$ which are standard normal
(due to the scaling with the standard deviation of a Brownian bridge) and have
a simple correlation structure: $\sqrt{s (1 - t)}/\sqrt{t (1 - s)}$ for
$s \le t$ and both $\in \{t_1, \dots, t_{m-1}\}$. Therefore, critical
values and $p$-values can be easily computed from a multivariate normal
distribution with standard normal marginals and this particular correlation
matrix; see also \citeNP{hotzei08} for more details. In R, this can be
accomplished using the mvtnorm package \cite{mvtnorm}.

@@

Simulation results are presented in Figure~\ref{fig:sim1res}.  Rows of
the figure 
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_o$ statistic
from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see Merkle and Zeileis, 2012, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than 0.05 at $d=0$).  This is because the likelihood
ratio test requires estimation of a multiple-groups model, which is very
unstable with large numbers of groups and small sample sizes as only
$n/m$ observations are available in each subsample.  The proposed
statistics are all of the LM-type and just require estimation of the single-group model,
leading to a clear advantage in these conditions.

@@

The more interesting result of this simulation lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the result 
of Bentler and Bonett \citeyear{Ben80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\mathit{WDM}_o$ statistic being the lowest and the $\max\mathit{LM}_o$
statistic being higher at larger values of $d$.  
This demonstrates that the proposed ordinal test
statistics can be especially useful at large sample sizes, where
traditional test statistics result in frequent significance.
Both statistics exhibited much lower power to detect a measurement
invariance violation that occurs only at a single level of $V$.

@@

With the positive psychology movement, the construct of gratitude has
received much research attention (for a review, see \citeNP{EmmMcc04}).
Recently, scholars have begun to explore gratitude
in youth. One potential problem with this research is that
scholars, with no exception, have used adult gratitude inventories to
measure youth gratitude, thus raising the question of whether the
existing gratitude scales used with adults are valid in research with
youth. Addressing this issue, \citeNP{FroFan11} had a
large sample of 
youth ($n$ = 1401, ranging from late childhood (10 years old) to
late adolescent (19 years old)) complete the three most widely used
adult gratitude inventories, including Gratitude Questionnaire~6
(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective
Checklist (GAC; \citeNP{MccEmm02}), and Gratitude, Resentment,
Appreciation Test-Short Form (GRAT-Short Form;
\citeNP{ThoWat03}). The authors were interested in whether the
youth factor structure 
for the gratitude scales resembles that of 
adults, and whether the gratitude scales are invariant across the
youth age groups.

@@

\citeNP{FroFan11} used confirmatory factor 
models to study the invariance of three youth gratitude scales across
students aged 10 to 19 years.  Due to sample size constraints, the age
variable included six categories: 10--11 years, 12--13 years, 14 years,
15 years, 16 years, and 17--19 years.  Thus, age is an ordinal variable
with which the proposed tests can be applied.

To test for measurement invariance w.r.t.\ age, each of the three scales
was individually 
factor-analyzed using the items that comprised the scale.
For each model, the authors 
first fit a congeneric model (all parameters free for each level of age),
followed by a tau-equivalent model (factor loadings restricted to be
equal across each level of age) and a parallel model (all parameters restricted
to be equal across levels of age).  Because their sample size was large ($n
\approx 1400$), they could not rely solely on likelihood ratio tests
(i.e., $\chi^2$ difference tests) for model comparison because the
tests were overly sensitive at their sample size.  To supplement these
tests, \citeNP{FroFan11} examined a set of alternative fit indices, including
the non-normed fit index, the comparative fit index, and the
incremental fit index. %% TODO cite
The authors generally found support for the tau-equivalent models
through these alternative fit indices; the likelihood ratio test often
resulted in significance even when the alternative indices indicated
good fit.

In the analyses described below, we re-analyze the \citeNP{FroFan11} data using
the ordinal test statistics proposed in this paper.  This results in a
series of tests that are less sensitive than the likelihood ratio test
to minor parameter discrepancies,
while being more sensitive to ordinal violations of measurement invariance.
We focus on two analyses from Froh et al.\ where the likelihood
ratio test resulted in significance (indicating that the restricted
model did not fit as well as the less-restricted model) but the
alternative fit measures indicated the opposite.
These include
comparison of a one-factor congeneric model to a one-factor
tau-equivalent model using the 
GQ6 and comparison of a one-factor tau-equivalent model to a one-factor
parallel model 
using the GAC.  To conduct equivalent analyses via the proposed tests, we
fit the more-restricted model in each case and test for instability in
the focal model parameters.

Of the 1401 cases originally collected by \citeNP{FroFan11}, we use here all
subjects with complete data (resulting in $n=1327$).

@@

<<gq6>>=
m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
          data=yg, group="agegroup", group.equal="loadings",
          meanstructure=TRUE)
gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, 
              order.by = as.numeric(yg$agegroup), 
              vcov = info_full, sandwich = FALSE, parm=1:4)
if(file.exists("ol2bb1.rda")){
  load("ol2bb1.rda")
} else {
  ol2bb1.fun <- ordL2BB(gefp1)
  save(ol2bb1.fun, file="ol2bb1.rda")
}
omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
ol2bb1 <- sctest(gefp1, functional=ol2bb1.fun)

@@

\subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
model to the GQ-6 data, Froh et al.\ used alternative fit indices to
conclude that the tau-equivalent model was as good as the congeneric
model.  However, the likelihood ratio test comparing these two models
was significant ($\chi^2_{20} = \Sexpr{round(lrt1[[5]][2], digits = 2)}, p=\Sexpr{round(lrt1[[7]][2], digits = 3)}$ for the data considered
here).

\fixme{Here and also subsequently, formulations like ``we compare model~A
and model~B'' are used. I'm not sure whether this sends the correct message.
If I want to compare exactly these two models, then the LRT is the correct
thing to do. But we are looking for other types of deviations, especially if
a split into two (rather than all) groups might be sufficient to capture
the effect of $V$. Hence, maybe we should explain in some more detail what
we are testing and also explore other models when rejecting the model? For
more comments in this vein see below.}
\readme{Achim: Added two sentences here that hopefully address the issue.}
We can use the proposed ordinal statistics to assess whether or not
the factor loadings in the tau-equivalent model fluctuate with respect
to age.  Unlike the LRT, the test does not require parameters to
differ within each subgroup.  Instead, we test for deviations such
that a split into two subgroups is sufficient to capture the effect of
$V$.
In employing the ordinal tests, we obtain~$\mathit{WDM}_o=\Sexpr{round(omax1$statistic, digits = 2)}, p=\Sexpr{round(omax1$p.value, digits = 3)}$
 and~$\max\mathit{LM}_o=\Sexpr{round(ol2bb1$statistic, digits = 2)}, p=\Sexpr{round(ol2bb1$p.value, digits = 3)}$.
Both $p$-values are clearly larger than that of the likelihood ratio test
and neither is significant at an $\alpha$ of 0.05,
which supports the conclusions that Froh et al.\ obtained from
alternative fit statistics.  This provides further evidence that 
there is no systematic deviation of the factor loadings along age
and that the likelihood ratio statistic is overly sensitive, picking
up some non-systematic dependence on age.

@@

<<gac>>=
m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, 
              order.by = as.numeric(yg$agegroup), 
              vcov = info_full, sandwich = FALSE, parm=3:6)
if(file.exists("ol2bb2.rda")){
  load("ol2bb2.rda")
} else {
  ol2bb2.fun <- ordL2BB(gefp2)
  save(ol2bb2.fun, file="ol2bb2.rda")
}
omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
ol2bb2 <- sctest(gefp2, functional=ol2bb2.fun)

@@

\subsubsection{GAC} In fitting tau-equivalent and parallel models to the
GAC data, \citeNP{FroFan11} obtained mixed results.  The alternative fit
indices did not all agree with one another, and the likelihood ratio test
indicated that the parallel model fit worse than the tau-equivalent
model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
for the data considered here).  Froh et al.\ ultimately 
concluded that the tau-equivalent model provided a better fit than did
the parallel model.

@@

Plots representing the statistics' fluctuations across age
groups are displayed in Figure~\ref{fig:gacres}.  The left panel
displays the process associated with $\mathit{WDM}_o$, while the right
panel displays the process associated with $\max\mathit{LM}_o$.  It is
observed that both processes are fully above the critical value,
implying the measurement invariance violation.  Additionally, both
processes peak at the 12--13 age group, suggesting that
parameters differ between individuals up to 13 years of age and older
than 13 years. \fixme{Maybe we could additionally fit a tau-equivalent
model for only two age groups: $\le 13$ vs.\ $> 13$. Then, we could
re-run tests and check whether this already fits sufficiently well?
This could help making the split into two subgroups clearer, rather
than always splitting into all possible age groups.}
\readme{Achim: See code starting on line 963 of the Rnw file. Even if we
  split around age 13, the test statistics are all significant.}

@@

The results presented in this paper imply that the proposed
ordinal statistics may ``miss'' measurement invariance violations that
are not monotonic w.r.t.\ $V$. More precisely, while the suggested tests
are also consistent for such non-monotonic violations, they seem to be
 less powerful than the likelihood ratio test.
We speculate that, in most applications, this will not be a major issue
because the researcher's a priori hypotheses exclusively focus on
monotonic measurement 
invariance violations.  For example, in the youth gratitude
application, we tested for measurement invariance across six age
groups.  If we observed a measurement invariance violation whereby
factor loadings were equal at all age groups except 14 years,
we would
have a hard time explaining the violation as anything but an anomaly
in the 14-year-olds.  Further, if $n$ is large, we are likely to
suspect that the result arises from the large sample size.
We may still be interested in why the
14-year-olds differed, but the analysis is purely exploratory at this
point because this type of violation was unexpected.  
However, there is generally a tradeoff between the ordinal statistics
and the likelihood ratio statistic.  The
proposed ordinal statistics usually provide more powerful tests of one's a
priori hypothesis regarding measurement invariance w.r.t.\ ordinal
$V$, while the likelihood ratio statistic provides a more powerful test of
general (non-monotonic) measurement invariance w.r.t.\ $V$. \readme{I
replace ``sharper'' by ``more powerful'' which would be the correct
statistical jargon. Or is sharper the terminus technicus in psychometrics?}
\readme{Achim: No, ``sharper'' was simply my terminus technicus when I
  was writing this section.}
While the
latter feature may be important in some high-stakes applications, many
researchers are likely to find the former feature appealing for their
work.

Along with using likelihood ratio tests to study measurement
invariance, researchers may wish to treat ordinal $V$ as continuous
(especially if $V$ has many levels).  
\fixme{We should also point to the supLM test of our first paper here.
The maxLM and supLM will generally be very close if the number of
potential thresholds is large. Maybe we could also re-iterate that
the forming of age groups a priori might not even be necessary and
more studies could collect data about age as a continuous rather than
just an ordinal variable.}
\readme{Achim: You say both ``maxLM'' and ``supLM'' above, but I assume
  you are generally referring to maxLM for ordinal $V$ vs maxLM for
  continuous $V$.  See sentences below.}
As described in detail by Merkle and Zeileis \citeyear{MerZei12}, we
can also use cumulative score processes with continuous $V$,
resulting in, e.g., maximum LM statistics and Cram\'{e}r-von Mises
statistics.  In fact, when the number of potential thresholds is
large, the proposed~$\max\mathit{LM}_o$ statistic will be very close
to the $\max\mathit{LM}$ statistic described in Merkle and Zeileis.
Thus, the formation of ordinal age groups (or other variables) is not
necessary for testing measurement invariance, and it may be beneficial
to collect continuous age data (e.g., measured by days instead of by 
years).

@@

\fixme{Should we make some practical remarks about the computation of
$p$-values somewhere? The issue with the maxLM test is that it is hard
to tabulate critical values in advance and hence simulation has to be
done for each case and is potentially rather time-consuming. Not sure
whether we should do this now or wait until the revision or so\dots}
\readme{Achim: I am happy to wait for the revision, and for now have
  inserted a TODO comment into the Rnw file.}

% TODO Practical remarks about the computation of p-values

>>>>

This section contains background on the theory underlying the
proposed statistics; for a more detailed
account, see \citeA{MerZei12}.

@@

As mentioned previously, when $V$ is categorical with a
relatively-small number of categories, 
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models usually cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals and then
examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test and are
discussed in detail by \citeA{MerZei12}. Here,
we review the tests' properties that are relevant for extending them
to the ordinal case.

@@

Test statistics that exhibit better performance in such situations 
aggregate information across parameters and possibly also across
individuals.  These test statistics include
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
with the former being a Cram{\'e}r-von Mises statistic and the latter
corresponding to a ``maximum'' Lagrange multiplier test, where the
maximum is taken across all possible divisions of individuals into two
groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
scaled by the asymptotic variance $t (1 - t)$ of the process
${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei12} found 
that both tests perform well when assessing simultaneous changes
in multiple factor analysis parameters, with the
$\mathit{CvM}$ test being somewhat advantageous in their particular
simulation setup. These simulations included 
situations in which subsets of model parameters were tested; such
situations are handled 
by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
that correspond to the parameters of interest.

@@

For the $\mathit{WDM}_o$ statistic, the resulting asymptotic distribution is
$\max_{j = 1, \dots, k} \max_{\ell = 1, \dots, m-1} {\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$.
Note that the effect of the outer maximum can be easily captured by a Bonferroni
correction as the $k$~components of the Brownian bridge are asymptotically independent.
Moreover, the inner maximum is taken over $m-1$ variables
${\bm B}^{0}(t_\ell) / \sqrt{t_\ell (1 - t_\ell)}$ which are standard normal
(due to the scaling with the standard deviation of a Brownian bridge) and have
a simple correlation structure: $\sqrt{s (1 - t)}/\sqrt{t (1 - s)}$ for
$s \le t$ and both $\in \{t_1, \dots, t_{m-1}\}$. Therefore, critical
values and $p$-values can be easily computed from a multivariate normal
distribution with standard normal marginals and this particular correlation
matrix; see also \citeA{hotzei08} for more details. In R, this can be
accomplished using the mvtnorm package \cite{mvtnorm}.

@@

Simulation results are presented in Figure~\ref{fig:sim1res}.  Rows of
the figure 
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_o$ statistic
from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see \citeNP{MerZei12}, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than 0.05 at $d=0$).  This is because the likelihood
ratio test requires estimation of a multiple-groups model, which is very
unstable with large numbers of groups and small sample sizes as only
$n/m$ observations are available in each subsample.  The proposed
statistics are all of the LM-type and just require estimation of the single-group model,
leading to a clear advantage in these conditions.

@@

The more interesting result of this simulation lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the result 
of \citeA{Ben80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\mathit{WDM}_o$ statistic being the lowest and the $\max\mathit{LM}_o$
statistic being higher at larger values of $d$.  
This demonstrates that the proposed ordinal test
statistics can be especially useful at large sample sizes, where
traditional test statistics result in frequent significance.
Both statistics exhibited much lower power to detect a measurement
invariance violation that occurs only at a single level of $V$.

@@

With the positive psychology movement, the construct of gratitude has
received much research attention (for a review, see \citeNP{EmmMcc04}).
Recently, scholars have begun to explore gratitude
in youth. One potential problem with this research is that
scholars, with no exception, have used adult gratitude inventories to
measure youth gratitude, thus raising the question of whether the
existing gratitude scales used with adults are valid in research with
youth. Addressing this issue, \citeA{FroFan11} had a
large sample of 
youth ($n$ = 1401, ranging from late childhood (10 years old) to
late adolescent (19 years old)) complete the three most widely used
adult gratitude inventories, including Gratitude Questionnaire~6
(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective
Checklist (GAC; \citeNP{MccEmm02}), and Gratitude, Resentment,
Appreciation Test-Short Form (GRAT-Short Form;
\citeNP{ThoWat03}). The authors were interested in whether the
youth factor structure 
for the gratitude scales resembles that of 
adults, and whether the gratitude scales are invariant across the
youth age groups.

@@

\citeA{FroFan11} used confirmatory factor 
models to study the invariance of three youth gratitude scales across
students aged 10 to 19 years.  Due to sample size constraints, the age
variable included six categories: 10--11 years, 12--13 years, 14 years,
15 years, 16 years, and 17--19 years.  Thus, age is an ordinal variable
with which the proposed tests can be applied.

To test for measurement invariance w.r.t.\ age, each of the three scales
was individually 
factor-analyzed using the items that comprised the scale.
For each model, the authors 
first fit a congeneric model (all parameters free for each level of age),
followed by a tau-equivalent model (factor loadings restricted to be
equal across each level of age) and a parallel model (all parameters restricted
to be equal across levels of age).  Because their sample size was large ($n
\approx 1400$), they could not rely solely on likelihood ratio tests
(i.e., $\chi^2$ difference tests) for model comparison because the
tests were overly sensitive at their sample size.  To supplement these
tests, \citeA{FroFan11} examined a set of alternative fit indices, including
the non-normed fit index, the comparative fit index, and the
incremental fit index. %% TODO cite
The authors generally found support for the tau-equivalent models
through these alternative fit indices; the likelihood ratio test often
resulted in significance even when the alternative indices indicated
good fit.

In the analyses described below, we re-analyze the \citeA{FroFan11} data using
the ordinal test statistics proposed in this paper.  This results in a
series of tests that are less sensitive than the likelihood ratio test
to minor parameter discrepancies,
while being more sensitive to ordinal violations of measurement invariance.
We focus on two analyses from \citeA{FroFan11} where the likelihood
ratio test resulted in significance (indicating that the restricted
model did not fit as well as the less-restricted model) but the
alternative fit measures indicated the opposite.
These include
comparison of a one-factor congeneric model to a one-factor
tau-equivalent model using the 
GQ6 and comparison of a one-factor tau-equivalent model to a one-factor
parallel model 
using the GAC.  To conduct equivalent analyses via the proposed tests, we
fit the more-restricted model in each case and test for instability in
the focal model parameters.

Of the 1401 cases originally collected by \citeA{FroFan11}, we use here all
subjects with complete data (resulting in $n=1327$).

@@

<<gq6>>=
m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
          data=yg, group="agegroup", group.equal="loadings",
          meanstructure=TRUE)
gefp1 <- gefp(m1, fit = NULL, order.by = as.numeric(yg$agegroup), 
              vcov = info_full, sandwich = FALSE, parm=1:4)
if(file.exists("ol2bb1.rda")){
  load("ol2bb1.rda")
} else {
  ol2bb1.fun <- ordL2BB(gefp1)
  save(ol2bb1.fun, file="ol2bb1.rda")
}
omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
ol2bb1 <- sctest(gefp1, functional=ol2bb1.fun)

@@

\subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
model to the GQ-6 data, \citeA{FroFan11} used alternative fit indices to
conclude that the tau-equivalent model was as good as the congeneric
model.  However, the likelihood ratio test comparing these two models
was significant ($\chi^2_{20} = \Sexpr{round(lrt1[[5]][2], digits = 2)}, p=\Sexpr{round(lrt1[[7]][2], digits = 3)}$ for the data considered
here).

We can use the proposed ordinal statistics to assess whether or not
the factor loadings in the tau-equivalent model fluctuate with respect
to age.  Unlike the LRT, the test does not require parameters to
differ across \emph{all} subgroups.  Instead, we test for deviations such
that a split into two subgroups is sufficient to capture the effect of $V$.
In employing the ordinal tests, we obtain~$\mathit{WDM}_o=\Sexpr{round(omax1$statistic, digits = 2)},
p = \Sexpr{format(round(omax1$p.value, digits = 3), nsmall = 3)}$
and~$\max\mathit{LM}_o=\Sexpr{round(ol2bb1$statistic, digits = 2)},
p=\Sexpr{round(ol2bb1$p.value, digits = 3)}$.
Both $p$-values are clearly larger than that of the likelihood ratio test
and neither is significant at $\alpha = 0.05$,
which supports the conclusions that \citeA{FroFan11} obtained from
alternative fit statistics.  This provides further evidence that 
there is no systematic deviation of the factor loadings along age
and that the likelihood ratio statistic is overly sensitive, picking
up some non-systematic dependence on age.

@@

<<gac>>=
m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp2 <- gefp(m2, fit = NULL, order.by = as.numeric(yg$agegroup), 
              vcov = info_full, sandwich = FALSE, parm=3:6)
if(file.exists("ol2bb2.rda")){
  load("ol2bb2.rda")
} else {
  ol2bb2.fun <- ordL2BB(gefp2)
  save(ol2bb2.fun, file="ol2bb2.rda")
}
omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
ol2bb2 <- sctest(gefp2, functional=ol2bb2.fun)

@@

\subsubsection{GAC} In fitting tau-equivalent and parallel models to the
GAC data, \citeA{FroFan11} obtained mixed results.  The alternative fit
indices did not all agree with one another, and the likelihood ratio test
indicated that the parallel model fit worse than the tau-equivalent
model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2], digits = 2)}, p<0.01$
for the data considered here).  \citeA{FroFan11} ultimately 
concluded that the tau-equivalent model provided a better fit than did
the parallel model.

@@

Plots representing the statistics' fluctuations across age
groups are displayed in Figure~\ref{fig:gacres}.  The left panel
displays the process associated with $\mathit{WDM}_o$, while the right
panel displays the process associated with $\max\mathit{LM}_o$.  It is
observed that both processes are fully above the critical value,
implying the measurement invariance violation.  Additionally, both
processes peak at the 12--13 age group, suggesting that
parameters differ between individuals up to 13 years of age and older
than 13 years. \fixme{Z: Maybe we could additionally fit a tau-equivalent
model for only two age groups: $\le 13$ vs.\ $> 13$. Then, we could
re-run tests and check whether this already fits sufficiently well?
This could help making the split into two subgroups clearer, rather
than always splitting into all possible age groups.}
\readme{ECM: See code starting on line 963 of the Rnw file. Even if we
  split around age 13, the test statistics are all significant.}
\readme{Z: My formulation above was misleading. We started out from
model m2 with 24 parameters where variances are restricted to be equal
across all groups. The LRT compares this to model m2.f with 44 parameters,
where the group-specific variances for all 6 age groups are estimated.
This ``costs'' us $(6 - 1) \cdot 4 = 20$ additional parameters. Instead,
we could fit a model m2.2 with 4 additional parameters -- i.e., 28 in total --
where the variances are just estimated separately for two age groups (split at age 14).
This should improve significantly upon m2 but if we are lucky it is not
significantly worse than m2.f.}

@@

The results presented in this paper imply that the proposed
ordinal statistics may ``miss'' measurement invariance violations that
are not monotonic w.r.t.\ $V$. More precisely, while the suggested tests
are also consistent for such non-monotonic violations, they seem to be
 less powerful than the likelihood ratio test.
We speculate that, in most applications, this will not be a major issue
because the researcher's a priori hypotheses exclusively focus on
monotonic measurement 
invariance violations.  For example, in the youth gratitude
application, we tested for measurement invariance across six age
groups.  If we observed a measurement invariance violation whereby
factor loadings were equal at all age groups except 14 years,
we would
have a hard time explaining the violation as anything but an anomaly
in the 14-year-olds.  Further, if $n$ is large, we are likely to
suspect that the result arises from the large sample size.
We may still be interested in why the
14-year-olds differed, but the analysis is purely exploratory at this
point because this type of violation was unexpected.  
However, there is generally a tradeoff between the ordinal statistics
and the likelihood ratio statistic.  The
proposed ordinal statistics usually provide more powerful tests of one's a
priori hypothesis regarding measurement invariance w.r.t.\ ordinal
$V$, while the likelihood ratio statistic provides a more powerful test of
general (non-monotonic) measurement invariance w.r.t.\ $V$. While the
latter feature may be important in some high-stakes applications, many
researchers are likely to find the former feature appealing for their
work.

Along with using likelihood ratio tests to study measurement
invariance, researchers may wish to treat ordinal $V$ as continuous
(especially if $V$ has very many levels).  
As described in detail by \citeA{MerZei12}, we
can also use cumulative score processes with continuous $V$,
resulting in, e.g., maximum LM statistics and Cram\'{e}r-von Mises
statistics.  In fact, when the number of potential thresholds is
large, the proposed~$\max\mathit{LM}_o$ statistic will be very close
to the $\max\mathit{LM}$ statistic described in \citeA{MerZei12}.
Thus, the formation of ordinal age groups (or other variables) is not
necessary for testing measurement invariance, and it may be beneficial
to collect continuous age data (e.g., age measured in days rather than in
years).

@@

% TODO % \fixme{Z: Should we make some practical remarks about the computation of
% TODO % $p$-values somewhere? The issue with the maxLM test is that it is hard
% TODO % to tabulate critical values in advance and hence simulation has to be
% TODO % done for each case and is potentially rather time-consuming. Not sure
% TODO % whether we should do this now or wait until the revision or so\dots}
% TODO % \readme{ECM: I am happy to wait for the revision, and for now have
% TODO %   inserted a TODO comment into the Rnw file.}
